## STATS 763: Assignment 2: Spinal surgery and readmission
#### Tarin Eccleston
#### 23/08/2024

We are interested in the causal effect of the duration of the surgery (Surgery time) on the probability of being readmitted within 30 days. The variable Re-admissions indicates the number of re-admissions within this period.

Assume that confounding is reasonably well alleviated by adjusting for Ethnicity, Trauma, Surgical complexity and Charlson Index (co-morbidity index) in models relating Surgery time to readmission within 30 days.

```{r results='hide', warning=FALSE, message=FALSE, error=FALSE, echo=FALSE}
library(tidyverse)
library(stringr)
library(lubridate)
```

### Causal Diagram

INSERT CAUSAL DIAGRAM HERE

### Data Manipulation

Based on the brief, we would like these columns in our dataframe for modelling:

- Re-admission
- Surgery time
- Ethnicity
- Trauma
- Surgical complexity
- Charlson Index (co-morbidity index)

#### Read Data

```{r results='hide', warning=FALSE, message=FALSE, error=FALSE}
spinal_surg <- read.csv("data/spinal_surgery.csv")
```


### Exploratory Analysis

### Modelling

### Discussion