## STATS 763: Assignment 3: Is is worth throwing fish back?
#### Tarin Eccleston
#### 25/09/2024

The file `hopper.csv` comes from a study of the survival of under-sized fish that are caught by beam trawler commercial fishing boats and returned to the sea.

These fish live on or near the sea bottom.  The study explored effects of observed environmental conditions at sea and of the randomly-assigned use of a water-filled storage hopper on the survival of the fish.

During trips with commercial trawlers, whole catches were unloaded from the  net  into either water-filled hoppers or conventional dry hoppers before sorting for size and species. For both hopper types, undersized fish were sampled from the sorting belt. After assessment of vitality status, sampled fish were housed in dedicated survival monitoring tanks on board. Upon return in the harbour fish were transferred to the laboratory to monitor their survival for up to 18 days post-catch. 

The researchers recorded sea conditions such as wave height and water temperature, or obtained them from public data sources.   

```{r results='hide', warning=FALSE, message=FALSE, error=FALSE, echo=FALSE}
library(tidyverse)
library(broom)
library(dagitty)
set.seed(123)
```

### Data Manipulation

```{r results='hide', warning=FALSE, message=FALSE, error=FALSE, echo=FALSE}
hopper <- read.csv("data/hopper.csv")
hopper <- hopper %>%
  rename(
    Wave_height = Wave_height..cm.,
    Depth = Depth..m.,
    Processing_time = Processing.time..min.,
    Haul_duration = Haul_duration..min.,
    Length = Length..cm.
  ) %>%
  mutate(Hopper = ifelse(Hopper == "Conventional", "Dry", Hopper))
```

### Exploratory Analysis

#### Causal Diagram

```{r}
```

#### Plots

```{r warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, class.output='centered-plot'}
ggplot(hopper, aes(fill=factor(Dead, levels = c(0, 1), labels = c("Survived", "Dead")), x=Hopper)) + 
    geom_bar(position="fill") +
    scale_fill_manual(values = c("Survived" = "#28A745", "Dead" = "#DC3545")) + 
    scale_y_continuous(labels = function(x) paste0(x * 100, "%")) +
    labs(x = "Hopper Type", y = "Percentage", title = "Percentage Survived and Dead Fish vs Hopper Type", fill = "Condition") +
    theme_minimal()
```

The average percentage of fish survival across all conditions seem to be less than 25%, with a slight increase in percentage of fish survival when fish are housed in a water-filled hopper. We must model with adjustments.

### 1. Which variables are potential confounders for the effect of hopper type on fish survival?

For a covariate to be a confounder of `Hopper` on fish survival (`Dead`), the cause (covariate) has to happen before the effect on hopper type. Since the study explored effects of the randomly-assigned use of water-filled storage hoppers or dry storage hoppers on the survival of the fish, there should be no confounders by design.

However, if we explored the effect of `Trip ID` on `Hopper`, the table shows that some trips have only dry hoppers. This could be due to the vessel not being compatible to have both a dry and water-filled hoppers. Certain trips will also have direct effects on fish survival probability. `Trip_ID` is therefore a confounder.

```{r warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, class.output='centered-plot'}
table(hopper$Trip_ID, hopper$Hopper)
```

### 2. Which variables are potential mediators of the effect of hopper type on fish survival?

For a covariate to be a mediator of `Hopper` on fish survival (`Dead`), `Hopper` needs to cause an effect on the covariate which in turn causes an effect on fish survival. Here are plausible mediators:

#### `Vitality Class`

```{r warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, class.output='centered-plot'}
ggplot(hopper, aes(fill=Hopper, x=Vitality_class)) + 
    geom_bar(position="fill") +
    scale_fill_manual(values = c("Dry" = "#D9A800", "Water" = "#007BFF")) +
    scale_y_continuous(labels = function(x) paste0(x * 100, "%")) +
    labs(x = "Vitality Class", y = "Percentage", title = "Percentage Water-filled and Dry Hopper vs Vitality Class", fill = "Hopper Type") +
    theme_minimal()
```

```{r warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, class.output='centered-plot'}
ggplot(hopper, aes(fill=factor(Dead, levels = c(0, 1), labels = c("Survived", "Dead")), x=Vitality_class)) + 
    geom_bar(position="fill") +
    scale_fill_manual(values = c("Survived" = "#28A745", "Dead" = "#DC3545")) + 
    scale_y_continuous(labels = function(x) paste0(x * 100, "%")) +
    labs(x = "Vitality Class", y = "Percentage", title = "Percentage Survived and Dead Fish vs Vitality Class", fill = "Condition") +
    theme_minimal()
```

The stacked barplots show whether the fish is put in a dry or in water-filled hopper should have an effect on it's vitality grade. The given vitality class also has an effect on the fish survival 18 days after being housed in a tank.

#### `Processing Time`

```{r warning=FALSE, message=FALSE, error=FALSE, echo=FALSE, class.output='centered-plot'}
medians <- hopper %>%
  group_by(Hopper) %>%
  summarize(median_time = median(Processing_time))

ggplot(hopper, aes(x = Processing_time, fill = factor(Hopper, labels = c("Dry", "Water Filled")))) + 
  geom_density(alpha = 0.8) +  # Density plots with transparency
  geom_vline(data = medians, aes(xintercept = median_time), 
             color = "black", linetype = "dashed", size = 1) +  # Add median lines
  geom_text(data = medians, aes(x = median_time, y = 0, 
                                label = paste("Median:", round(median_time, 2))),
            color = "black", vjust = -1.5, hjust = 1.1) +
  scale_fill_manual(values = c("Dry" = "#D9A800", "Water Filled" = "#007BFF")) +
  labs(x = "Processing Time (minutes)", y = "Density", title = "Processing Time by Hopper Type") +
  facet_wrap(~factor(Hopper, labels = c("Dry", "Water Filled")), scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
medians <- hopper %>%
  group_by(Dead) %>%
  summarize(median_time = median(Processing_time))

ggplot(hopper, aes(x = Processing_time, fill=factor(Dead, levels = c(0, 1), labels = c("Survived", "Dead")))) + 
  geom_density(alpha = 0.8) +  # Density plots with transparency
  geom_vline(data = medians, aes(xintercept = median_time), 
             color = "black", linetype = "dashed", size = 1) +  # Add median lines
  geom_text(data = medians, aes(x = median_time, y = 0, 
                                label = paste("Median:", round(median_time, 2))),
            color = "black", vjust = -1.5, hjust = 1.1) +
  scale_fill_manual(values = c("Survived" = "#28A745", "Dead" = "#DC3545")) +
  labs(x = "Processing Time (minutes)", y = "Density", title = "Processing Time by Survival Status") +
  facet_wrap(~factor(Dead, labels = c("Survived", "Dead")), scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none")
```

There appears to be a slight trend; however, we still include `Processing_time` as a mediator between `Hopper` and fish survival. The time it takes for a fish to move from onboarding to storage tanks varies depending on whether they are placed in dry or water-filled hoppers. On average, water-filled hoppers require slightly more time to process fish from the hopper to the tanks compared to dry hoppers. Longer processing times are expected to lead to lower fish survival rates, as demonstrated in the plot above.

### 3. Estimate the effect of hopper type on survival.

We make effort to exclude `Vitality_class` and `Processing_time`to explain the causal effects `Hopper` on fish survival as they are potential mediators for `Hopper` on fish survival. We don't need to include environmental conditions, haul duration or fish length as based on the causal diagram. We add `Trip_ID` into the model as that is the only confounder of `Hopper` on fish survival.

```{r}
fish_survival_model <- glm(Dead ~ Hopper + Trip_ID, family = binomial(link = "logit"), data = hopper) 
summary(fish_survival_model)
```

```{r}
est <- tidy(fish_survival_model)$estimate[2]
variance <- tidy(fish_survival_model)$std.error[2]^2
variance_delta = exp(2*est) * variance^2

ci_lower <- exp(est) - 1.96 * sqrt(variance_delta)
ci_upper <- exp(est) + 1.96 * sqrt(variance_delta)

ci <- c(ci_lower, exp(est), ci_upper)
names(ci) <- c("2.5%", "Estimate", "97.5%")
ci <- 1 - ci
ci
```

For a fish that was housed in a hopper type that was water-filled compared to dry, there is an expected decrease by 29.4% to 20.5% in odds of death after 18 days post-catch.

### 4. How does the effect of hopper type on survival vary with fish characteristics and sea conditions?

Introduce interaction term with hopper type and fish characteristics and sea conditions. Explore interaction terms one at a time as it could be too overwhelming to do all at once.

```{r}
fish_survival_hopper_temp_water_interaction_model <- glm(Dead ~ Trip_ID + Hopper*(Temp_water), family = binomial(link = "logit"), data = hopper) 
summary(fish_survival_hopper_temp_water_interaction_model)
```

```{r}
fish_survival_hopper_wave_height_interaction_model <- glm(Dead ~ Trip_ID + Hopper*(Wave_height), family = binomial(link = "logit"), data = hopper) 
summary(fish_survival_hopper_wave_height_interaction_model)
```

```{r}
fish_survival_hopper_depth_interaction_model <- glm(Dead ~ Trip_ID + Hopper*(Depth), family = binomial(link = "logit"), data = hopper) 
summary(fish_survival_hopper_depth_interaction_model)
```

```{r}
fish_survival_hopper_substrate_interaction_model <- glm(Dead ~ Trip_ID + Hopper*(Substrate), family = binomial(link = "logit"), data = hopper) 
summary(fish_survival_hopper_substrate_interaction_model)
```

```{r}
fish_survival_hopper_length_interaction_model <- glm(Dead ~ Trip_ID + Hopper*(Length), family = binomial(link = "logit"), data = hopper) 
summary(fish_survival_hopper_length_interaction_model)
```

For evaluating interaction terms, we look at the p-values for the coefficients of interest. Large p-values; greater than 0.05 for `Hopper:Temp_water`, `Hopper:Wave_height`, `Hopper:Depth`, `Hopper:Substrate`, and `Hopper:Length` (of fish) coefficients indicate there is insignificant evidence to suggest that estimates for `Hopper` vary with fish characteristics and sea conditions.

### 5. To what extent would measuring vitality class (which is rapid and simple) be a good substitute for measuring actual fish survival (which is slow and expensive)?

Fit a model for `Vitality_class` on fish survival. We include `Hopper` and `Trip_ID` as confounders.

```{r}
test_model <- glm(Dead ~ Vitality_class + Hopper + Trip_ID - 1, family = binomial(link = "logit"), data = hopper) 
summary(test_model)
```

High p-values for across the board for `Vitality_class` levels suggest a strong relationship between vitality class and fish survival, low p-values for `Hopper` and `Trip_ID` suggest weak direct relationship between these covariates and fish survival. 

- INSERT STACKED BAR CHART

Good predictor, high p-values

### Discussions

- Sampling based on length, possible confounder?
- putting them in a tank could be problematic
- In general fishing results in low survival rates of fish

