## STATS 763: Lab 4: Nonlinear functions of predictors
#### Tarin Eccleston
#### 10/08/2024

```{r results='hide', warning=FALSE, message=FALSE, error=FALSE, echo=FALSE}
library(tidyverse)
library(stringr)
library(lubridate)
```

### Quadratics
#### Q1) Simulate x <- rexp(100) and y <- 4*exp(-x)+rnorm(100). Plot y against x and overlay the true form of ð¸[ð‘Œ|ð‘‹=ð‘¥] and the fitted value from a linear model with x and x^2 as predictors. How would you describe the regression results to avoid giving the misleading impression that the relationship changes direction?

```{r}
x <- rexp(100)
y_expected <- 4*exp(-x)
y <- 4*exp(-x)+rnorm(100)
y_x = data.frame(cbind(x, y_expected, y))

y_model_x <- lm(y ~ x, y_x)
y_model_x_squared <- lm(y ~ x + I(x^2), y_x) # need to revise this

y_x$y_fitted_x <- predict(y_model_x, y_x)
y_x$y_fitted_x_squared <- predict(y_model_x_squared, y_x)
```

```{r}
ggplot(y_x, aes(x = x)) +
  geom_point(aes(y = y), color = "blue", alpha = 0.5) +
  geom_line(aes(y = y_expected), color = "green", size = 1) +
  geom_line(aes(y = y_fitted_x), color = "red", size = 1) +          
  geom_line(aes(y = y_fitted_x_squared), color = "purple", size = 1) + 
  labs(title = "True Curve vs Fitted Curves",
       x = "x",
       y = "y") +
  theme_minimal()
```


Support for ___ to ___. 
