# STATS 763: Assignment 1: Cycling in the Rain
### Tarin Eccleston
### 07/08/2024

```{r results='hide', warning=FALSE, message=FALSE, error=FALSE}
library(tidyverse)
```

#### Read Data

```{r results='hide', warning=FALSE, message=FALSE, error=FALSE}
ac_2018 <- read_csv("data/aucklandcyclists2018.csv")
ac_2019 <- read_csv("data/aucklandcyclists2019.csv")
ac_2020 <- read_csv("data/aucklandcyclists2020.csv")
ac_2021 <- read_csv("data/aucklandcyclists2021.csv")
ac_2022 <- read_csv("data/aucklandcyclists2022.csv")

monitoring_locations <- read.csv("data/monitoringlocations.csv")

daily_rainfall <- read.csv("data/daily-rainfall-at-30-sites-1960-to-2022.csv")
daily_rainfall_data_dict <- read.csv("data/daily-rainfall-at-30-sites-1960-to-2022-dictionary.csv")
```

#### Todo: Standardising Date Format

We need all date formats to be in YYYY-MM-DD.

```{r}

```

#### Filter out important rainfall data

Two of the rainfall monitoring locations are relevant to Auckland: "Auckland (Auckland)", close to the airport, and "Whangaparāoa (Auckland)".

We only need need rainfall data corresponding to cycling data between 2018-2022.

```{r}
daily_rainfall <- daily_rainfall %>%
  filter(site == "Auckland (Auckland)" | site == "Whangaparāoa (Auckland)") %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d")) %>%
  filter(date >= as.Date("2018-01-01"))
```


